A hazards approach to the biometric analysis of infant mortality

### 2023-03-21 paper2.Rmd
editorial

The cumulative numbers of death are plotted against the
corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
after 28 days (68 on the $g$ scale) is excellent for both time periods.

@@

Instead of using the total number of births throughout in the denominator as in  the
Bourgeois-Pichat procedure, we suggest 
using the present risk-set size, that is, the the total number of infants still alive 
immediately prior to the death age under investigation. Given proper data, it is easily achieved.
For instance,by the use of 
the function `risksets` in the **R** [@cran] package `eha` [@eha; @ehar2] we can get Table \@ref(tab:risksets).

@@

The plot corresponding to Figure \@ref(fig:first) is shown in Figure \@ref(fig:second).

@@

We note that the Bourgeois-Pichat method requires that no right censorings (infants lost 
to follow-up before one year of age) are present, and no left truncations (infants under observation 
only from an age later than birth). The hazards approach, on the other hand, allows
left truncation and right censoring as long as they are *non-informative* in the usual sense. Often, though,
this not very important, since new-born babies tend to be geographically stationary
for their first year of life.

Throughout the rest of the paper, everything happens on the $g$ time scale.

@@

 We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).

In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 28.

@@

We now consider the whole infant age span (on the $g$ scale), and compare the uniform and exponential densities, especially of interest is the fraction of deaths that occur in the neonatal phase. @manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant, 
18.56 percent, independent of the overall level of infant mortality $p$. This is not the case for the hazards model, see Table \@ref(tab:exog).

```{r exog}
source("R/tbl.R")
p <- c(0.005, 0.01, 0.05, 0.1, 0.2)
lambda <- -log(1 - p) / 365
n <- length(p)
exo28 <- numeric(n)
exo365 <- numeric(n)
for (i in 1:n){
    exo28[i] <- pexp(g(28), lambda[i]) * 1000
    exo365[i] <- pexp(g(365), lambda[i]) * 1000
}
utt <- cbind(exo28, exo365, round(100 * (exo28 / exo365), 2))
utt <- round(utt, 3)
colnames(utt) <- c("Neonatal", "Total", "Fraction")
tbl(utt, caption = "Distribution over age periods of exogenous mortality (per 1000).")
```

@@

We see that the post-neonatal part of the total cumulative hazards are more or
less convex (bend upward at the end) for those born in spring (red) or summer (green), while 
the curves are concave for births in the winter (blue). Bad fits to "constant hazard"
and "uniform distribution". However, as the solid black curve shows the 12-month mixture 
has an excellent fit to a straight line.

@@

The assumptions about constant hazard and uniform distribution of exogenous mortality 
are only applicable on population level, it can never hold for an individual. It can 
be thought of as an average over birth cohorts, where it is important to have birth dates 
spread evenly within calendar years.

@@

The crude IMR in Umeå 1891--1920 was around `r crudprob` per thousand, clearly higher than 
the later time period. Let us perform the biometric analysis with these data, see 
Figure \@ref(fig:bioume9120).

@@

However, this data set is known to be of very high quality [@tbcl99], so we are inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.

@@

An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
There is no room for endogenous deaths at all, but otherwise a reasonably good fit with both 
methods.

@@

>>>>

The cumulative numbers of deaths, divided by *the number of births*, are plotted against the
corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
after 28 days (68 on the $g$ scale) is excellent for both time periods.

@@

Instead of using the total number of births throughout in the denominator as in  the
Bourgeois-Pichat procedure, we suggest 
using the present risk-set size, that is, the the total number of infants still alive 
immediately prior to the death age under investigation. Given proper data, it is easily achieved.
For instance, by using
the function `risksets` in the **R** [@cran] package `eha` [@eha; @ehar2] we get Table \@ref(tab:risksets).

@@

The hazard-based plot corresponding to Figure \@ref(fig:first) is shown in Figure \@ref(fig:second).

@@

We note that the Bourgeois-Pichat method requires that no right censorings (infants lost 
to follow-up before one year of age) are present, and no left truncations (infants under observation 
only from an age later than birth). The hazards approach, on the other hand, allows
left truncation and right censoring. Often, though,
this not very important, since new-born babies tend to be geographically stationary
for their first year of life.

@@

 We start with the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).

In order to investigate this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 28.

@@

We now consider the whole infant age span (on the $g$ scale), and compare the uniform and exponential densities, especially of interest is the fraction of deaths that occur in the neonatal phase. @manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant, 
18.56 percent, independent of the overall level of infant mortality $p$. This is not the case for the hazards model, see Table \@ref(tab:exog).
As the total mortality increases, the fraction exogenous deaths in the neonatal period also increases.
```{r exog}
source("R/tbl.R")
p <- c(0.005, 0.01, 0.05, 0.1, 0.2)
lambda <- -log(1 - p) / 365
n <- length(p)
exo28 <- numeric(n)
exo365 <- numeric(n)
for (i in 1:n){
    exo28[i] <- pexp(g(28), lambda[i]) * 1000
    exo365[i] <- pexp(g(365), lambda[i]) * 1000
}
utt <- cbind(exo28, exo365, round(100 * (exo28 / exo365), 2))
utt <- round(utt, 3)
colnames(utt) <- c("Neonatal", "Total", "Fraction")
tbl(utt, caption = "Distribution over age periods of exogenous mortality (per 1000).")
```

@@

We see that the post-neonatal part of the total cumulative hazards are more or
less convex (bend upward at the end) for those born in spring (red) or summer (green), while 
the curves are concave for births in the winter (blue). Bad fits to "constant hazard"
and "uniform distribution". However, as the solid black curve shows, the 12-month mixture 
has an excellent fit to a straight line.

@@

The assumptions about constant hazard and uniform distribution of exogenous mortality 
are only applicable on the population level, it can never hold for an individual. It can 
be thought of as an average over birth cohorts, where it is important to have birth dates 
spread evenly within calendar years.

@@

The crude IMR in Västerbotten 1891--1920 was around `r crudprob` per thousand, clearly higher than 
the later time period. Let us perform the biometric analysis with these data, see 
Figure \@ref(fig:bioume9120).

@@

However, this data set is known to be of very high quality [@tbcl99], so we are inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
Specifically, the log-cube transformation, which fits well with 19th century data, seems
to be completely out of order in the eighteenth century.

@@

An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
There is no room for endogenous deaths at all, but otherwise a reasonably good fit with both 
methods in the post-neonatal period..

@@

On the other hand, the whole business with the 
biometric analysis and the log-cube transform seems to be highly questionable. 
It works for data in approximately the same time period and similar environment, 
but beyond that, it is not generalizable at all. These methods should be used with 
great care, if at all.
